---
title: "Greenhouse Emissions: Influence of Methane and CO₂"
author: "Saakshi Agarwal"
format: 
  pdf: 
    geometry:
      - top=18mm
      - left=18mm
      - right=18mm
      - bottom=22mm
      - heightrounded
    colorlinks: true
papersize: a4
fontsize: 11pt    
mainfont: Latin Modern Roman
sansfont: Latin Modern Roman
include-in-header: 
      text: |
        \usepackage{titling}
        \setlength{\droptitle}{-20mm}   % move title up
        \pretitle{\begin{center}\large\bfseries }
        \posttitle{\end{center}  \vspace*{-0.5cm}}        % set format of title
        \preauthor{\begin{center}}
        \postauthor{\end{center}\ignorespacesafterend \vspace*{-2cm}}
---

This analysis explores the relationship between total greenhouse gas emissions(GHG) (kt of CO2-eq) and key environmental contributors. Data from dataset wdi_2020 which have 217 observations, and CO2 emissions(tons per capita) and methane emissions(METH) (kt of CO2-eq) are chosen as independent variables due to their significant role in climate change.

```{r}
#| label: setup
#| echo: false
#| results: hide
#| message: false
#| warning: false
# install packages
library(knitr)
library(tidyverse)
library(readxl)
library(GGally)
library(tinytex)
library(dplyr)


# set default chunk options
# options can be overridden within individual chunks
opts_chunk$set(fig.align='center', 
               out.width="50%",
               message = FALSE,
               echo = FALSE,
               results = "hide",
               eval = TRUE)

# set options for printing
options(width = 70,
        pillar.print_min = 6,
        pillar.print_max = 10
        )


# update theme for all plots (optional)
theme_update(text = element_text(size = 16)
             #, plot.background = element_rect(colour = "grey", fill=NA, linewidth=0) # add border around plot to make it more visible
             , panel.border = element_rect(colour = "black", fill=NA, linewidth=1)
             , panel.background = element_rect(fill = "white")
           )
```

```{r, include=FALSE}
#| echo: false
#| label: read_data

#reading the csv file
data = read.csv("C:\\Users\\saaks\\Downloads\\wdi_2020.csv")

#renaming it for better variable name
wdi <- data %>%
  rename(
    Total_GDP = NY.GDP.MKTP.CD,
    GDP_Capita = NY.GDP.PCAP.CD,
    Working_Population = SP.POP.1564.TO.ZS,
    Children_Population = SP.POP.0014.TO.ZS,
    Elderly_Population = SP.POP.65UP.TO.ZS,
    Female_Population = SP.POP.TOTL.FE.ZS,
    Male_Population = SP.POP.TOTL.MA.ZS,
    Total_Population = SP.POP.TOTL,
    Poverty_Gap_Low_Income = SI.POV.GAPS,
    Poverty_Gap_Mid_Income = SI.POV.UMIC.GP,
    Measles_Immunization = SH.IMM.MEAS,
    Diphtheria_Immunization = SH.IMM.IDPT,
    Women_in_Parliament = SG.GEN.PARL.ZS,
    CO2_Emissions = EN.ATM.CO2E.PC,
    Methane = EN.ATM.METH.KT.CE,
    Greenhouse_Gas= EN.ATM.GHGT.KT.CE,
    Access_to_Clean_Water = SH.H2O.SMDW.ZS,
    Urban_Clean_Water_Access = SH.H2O.SMDW.UR.ZS,
    Rural_Basic_Water = SH.H2O.BASW.RU.ZS
  )
glimpse(wdi)

#checking columns and NA values
colSums(is.na(wdi))

#creating a new dataset for 3 variables
GHG =  wdi |> select(Greenhouse_Gas,CO2_Emissions,Methane)
GHG = na.omit(GHG)
#removing null values
colSums(is.na(GHG))

```

The raw data for GHG, CO2, and METH exhibited skewness, likely due to the wide range of values across different countries. Logarithmic transformation was applied to reduce heteroscedasticity and make the distribution closer to normal as shown in @fig-correlation.  A strong positive correlation can be observed between GHG and METH with r = 0.96 and a weak positive correlation with CO2 with r = 0.26. These correlations justify including CO2 and methane emissions as independent variables in the regression model.

```{r, echo=FALSE}
#| label: fig-correlation
#| fig-cap:  Data correlation and Distribution
#| fig.width: 6
#| fig.height: 4
#| out.width: 90%
#| fig-subcap:
#|   - "Original Data"
#|   - "Log Transformed Data" 
#| layout-ncol: 2

#checking correlation between them
C1 = cor(GHG$Greenhouse_Gas, GHG$Methane)
C2 = cor(GHG$Greenhouse_Gas, GHG$CO2_Emissions)
C3 = cor(GHG$CO2_Emissions, GHG$Methane)

ggpairs(GHG, lower = list(combo = wrap("facethist", binwidth = 5)))+
  theme(axis.text.x = element_text(size = 7),
       axis.text.y = element_text(size = 8),
       strip.text = element_text(size = 8))

#converting GHG, METH and CO2 in log values and combing them in a dataframe
GH2= GHG |> mutate(logGH = log10(GHG$Greenhouse_Gas),logME = log10(GHG$Methane), 
                   logCO = log10(GHG$CO2_Emissions) ) |> 
  select(logGH, logCO, logME)


#checking the correlation between independent and dependent variable
ggpairs(GH2)+theme(axis.text.x = element_text(size = 12),
                        axis.text.y = element_text(size = 8),   
                        strip.text = element_text(size = 8))
```

To predict GHG, a multiple regression model was used; a summary can be seen in Table 1. The equation is y = 0.82 + 0.31x1 + 0.92x2, indicating that an increase in 1 unit of CO2 can contribute upto a 31.6% rise in GHG, keeping METH constant. Similarly, an increase of 1 unit in METH can contribute upto 92.2% rise in GHG, keeping CO2 constant. The adjusted R\^2(0.9515) suggests that CO2 and METH are highly predictive of total GHG, with minimal unexplained variability. A high F-value suggests that the combined effect of CO2 and METH on total GHG is statistically significant, with a p-value \< 0.001 indicating that the model is highly significant.

```{r, echo=FALSE, include = FALSE}
#Running regression for the variables
GH_lm = lm(logGH~logCO+logME, data = GH2)
summary(GH_lm)
```

| Coefficient                               | Estimate  | Std_Error |
|-------------------------------------------|-----------|-----------|
| y = Intercept                             | 0.824     | 0.06      |
| x1 = Log CO2 per Capita (tons per capita) | 0.316     | 0.02      |
| x2 = Log METH (kt of CO2 eq)              | 0.921     | 0.01      |
| Adjusted R-square                         | 0.805     |           |
| F-Value                                   | 1863      |           |
| P-Value Model                             | p\< 0.001 |           |

: Summary of Multiple Linear Regression Model

Model assumptions were validated, as the residual vs. fitted plot does not show a pattern, suggesting homoscedasticity and the histogram of residuals shows a normal distribution, indicating that residuals are approximately normally distributed, as seen in @fig-residuals-plots. The scatter plot shows a random distribution of residuals, suggesting that CO2 and METH have a linear relationship with GHG, supporting the assumption of linearity, as seen in @fig-residuals-plots-variable. The collinearity between CO2 and METH is 0.06, which eliminates multicollinearity.

.

```{r, echo=FALSE}
#| label: fig-residuals-plots
#| fig-cap: "Residual plots for the multiple regression model" 
#| fig.height: 4
#| fig.width: 12
#| out-width: 90%

#Assumption checking
# Fitted values vs residuals
q1 <- ggplot(GH_lm, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(title='Residual vs. Fitted Values Plot',
       x='Fitted Values', y='Residuals')+theme_minimal()

# create new tibble with original data and residuals
GH_lm <- cbind(GH2, residuals = GH_lm$residuals)

# Distribution of residuals
q2 <- ggplot(GH_lm) +
  geom_histogram(aes(x = residuals), bins=5) +
  labs(x = "Residuals")+theme_minimal()

cowplot::plot_grid(q1, q2)+theme_minimal()
```

```{r, echo=FALSE}
#| label: fig-residuals-plots-variable
#| fig-cap: "Residual plots for each independent variable" 
#| fig.height: 4
#| fig.width: 12
#| out-width: 90%

# Independent Variable vs Residuals - plot
q3 <- ggplot(GH_lm, aes(x = logME , y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x= "Methane Emission (kt CO2 eq)",y='Residuals')+theme_minimal()

q4 <- ggplot(GH_lm, aes(x = logCO, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0) +
    labs(x= "CO2 emission per capita (tons per capita)",y='Residuals')+theme_minimal()

 cowplot::plot_grid( q3, q4)+theme_minimal()
```

In conclusion, these findings underscore the critical role of reducing methane and CO₂ emissions in mitigating climate change. Targeted policies addressing methane reduction, like waste management, fossil fuel extraction, and CO₂ reduction by using renewable energy, could significantly impact. These will reduce overall greenhouse gas levels and mitigate long-term environmental risks.
